---
title: "An aside on statistical summaries"
author: "Ryan Hill"
date: "23 March 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(datasauRus)
library(ggplot2)
library(gganimate)
library(dplyr)
```

## The Data

We have 13 datasets, each with 142 datapoints with an x and y coordinate. We might want to quickly analyse this data, say get the mean and standard deviation of each dataset?

```{r analyse}
datasaurus_dozen %>% 
  group_by(dataset) %>%
  summarise(mnx = mean(x), 
            sdx = sd(x),
            mny = mean(y), 
            sdy = sd(y))
```

Wow that's amazing! All of our datasets have the exact same mean and standard deviation (to 2 decimal places anyway). We can even add in the correlation to see how well correlated the x and y are for each dataset:


```{r analyse2}
datasaurus_dozen %>% 
  group_by(dataset) %>%
  summarise(cor = cor(x, y))
```

## Plotting the data

Okay so they seem pretty similar overall. But let's plot the data and see just how similar they are...

```{r pressure, echo=FALSE}
ggplot(datasaurus_dozen, aes(x = x, y = y, colour = dataset)) + 
  geom_point() + 
  facet_wrap(~dataset) +
  theme(legend.position = "none")
```

It turns out they aren't similar at all! All 13 of the datasets are completely different and show a completely different "relationship" to each other. This is why it's always useful and important to visualize and understand the data rather than just blindly follow the numbers.

```{r ani, echo = FALSE}
ggplot(datasaurus_dozen, aes(x=x, y=y))+
  geom_point()+
  theme_minimal() +
  transition_states(dataset, 3, 1) + 
  ease_aes('cubic-in-out')
```